---
title: "ETC5521: Exploratory Data Analysis"
subtitle: "Working with a single variable, making transformations, detecting outliers, using robust statistics"
author: "Di Cook"
email: "ETC5521.Clayton-x@monash.edu"
date: "Week 5 - Session 2"
length: "50 minutes"
titlebgimg: "images/bg-01.png"
output:
  xaringan::moon_reader:
    css:
      - ninjutsu 
      - "assets/font-awesome-all.css"
      - "assets/tachyons-addon.css"
      - "assets/animate.css"
      - "assets/fira-code.css"
      - "assets/boxes.css"
      - "assets/table.css"
      - "assets/styles.css"
      - "assets/monash-brand.css"
      - "assets/monash-fonts.css"
      - "assets/slide-types.css"
      - "assets/panelset-modified.css"
      - "assets/scroll.css"
    self_contained: false 
    seal: false 
    chakra: 'lib/remark-latest.min.js'
    lib_dir: lib
    includes:
      in_header: "assets/head.html"
    mathjax: "lib/mathjax-local/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    nature:
      highlightStyle: magula
      highlightLanguage: r 
      highlightLines: true
      highlightSpans: false 
      countIncrementalSlides: false
      slideNumberFormat: '%current%/%total%'
      navigation:
        scroll: false 
        touch: true
        click: false
      ratio: '16:9'
---

```{r, include = FALSE}
current_file <- knitr::current_input()
basename <- gsub(".Rmd$", "", current_file)
```

```{r, include = FALSE}
library(tidyverse)
library(colorspace)
library(patchwork)
options(width = 200)
knitr::opts_chunk$set(
  fig.path = "images/week4B/",
  fig.width = 6,
  fig.height = 3.5,
  fig.align = "center",
  dev.args = list(bg = "transparent"),
  # out.width = "100%",
  fig.retina = 3,
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  cache.path = "cache/week4B/"
)
theme_set(ggthemes::theme_gdocs(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "transparent", colour = NA), axis.line.y = element_line(color = "black", linetype = "solid"),
    plot.title.position = "plot",
    plot.title = element_text(size = 16),
    panel.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA)
  ))
```


```{r titleslide, child="assets/titleslide.Rmd"}
```


---



class: transition middle

# Categorical variables

<br><br>


This lecture is based on Chapter 4  of <br><br>Unwin (2015) Graphical Data Analysis with R

---

# There are two types of categorical variables

--

<br><br>

.monash-blue[**Nominal**] where there is no intrinsic ordering to the categories<br>
**E.g.** blue, grey, black, white.

--

<br>

.monash-blue[**Ordinal**] where there is a clear order to the categories.<Br>
**E.g.** Strongly disagree, disagree, neutral, agree, strongly agree.


---

# Categorical variables in R

.grid[
.item.br[
* In R, categorical variables may be encoded as **factors**.
.f4[
```{r factors, echo = TRUE}
data <- c(2, 2, 1, 1, 3, 3, 3, 1)
factor(data)
```
]
* You can easily change the labels of the variables:
.f4[
```{r factor-labels, echo = TRUE}
factor(data, labels = c("I", "II", "III"))
```
]
]
.item.f4[

{{content}}

]
]

--

* Order of the factors are determined by the input:


```{r factor-input, echo = TRUE}
# numerical input are ordered in increasing order #<<
factor(c(1, 3, 10))
# character input are ordered alphabetically #<<
factor(c("1", "3", "10"))
# you can specify order of levels explicitly #<<
factor(c("1", "3", "10"),
  levels = c("1", "3", "10")
)
```


---

# Numerical factors in R

```{r, echo = TRUE, warning = TRUE}
x <- factor(c(10, 20, 30, 10, 20))
mean(x)
```

--

<i class="fas fa-exclamation-triangle"></i> `as.numeric` function returns the internal integer values of the factor
```{r, echo = TRUE}
mean(as.numeric(x))
```

--

You probably want to use:

.flex[
.w-50[
```{r, echo = TRUE}
mean(as.numeric(levels(x)[x]))
```

]
.w-50[
```{r, echo = TRUE}
mean(as.numeric(as.character(x)))
```

]
]

---
# Numerical summaries: counts, proportions, percentages and odds

.flex[
.item.w-60[
.scroll-sign[.s500.f4[
```{r}
# https://www.who.int/teams/global-tuberculosis-programme/data
tb <- read_csv(here::here("data/TB_notifications_2023-08-21.csv"))
tb_tot <- tb %>% 
  rowwise() %>%
  mutate(count = sum(c(new_sp, new_sn, new_su, new_ep, new_oth), na.rm=T)) %>%
  select(country, iso3, year, count) %>%
  ungroup()
tb_oz <- tb_tot %>%
  filter(iso3 == "AUS")
```

```{r}
options(digits=2)
tb_oz %>%
  filter(year >= 2000) %>%
  mutate(p = count/sum(count),
         pct = p*100, 
         odds = count/count[year==2000]) %>%
  print(n=100)
```
]]
]
.item.w-40[

For qualitative data, compute 

- count/frequency, 
- proportion/percentage
- and sometimes, an **odds ratio**. Here we have used ratio relative to the count in year 2000. 

<br><br>
**Note:** For exploration, no rounding of digits was done, but to report you would need to make the numbers pretty.
]]
---

# .orange[Revisiting Case study] .circle.bg-orange.white[1] 2019 Australian Federal Election

```{r aus-election-data1}
df1 <- read_csv(here::here("data/HouseFirstPrefsByCandidateByVoteTypeDownload-24310.csv"),
  skip = 1,
  col_types = cols(
    .default = col_character(),
    OrdinaryVotes = col_double(),
    AbsentVotes = col_double(),
    ProvisionalVotes = col_double(),
    PrePollVotes = col_double(),
    PostalVotes = col_double(),
    TotalVotes = col_double(),
    Swing = col_double()
  )
)
```
```{r aus-election-data3}
tdf3 <- df1 %>%
  group_by(DivisionID) %>%
  summarise(
    DivisionNm = unique(DivisionNm),
    State = unique(StateAb),
    votes_GRN = TotalVotes[which(PartyAb == "GRN")],
    votes_total = sum(TotalVotes)
  ) %>%
  mutate(perc_GRN = votes_GRN / votes_total * 100)
```
.panelset[
.panel[.panel-name[📊]

.flex[
.w-50[
```{r aus-election-plot1, fig.height = 5}
tdf3 %>%
  ggplot(aes(perc_GRN, State)) +
  ggbeeswarm::geom_quasirandom(groupOnX = FALSE, varwidth = TRUE) +
  labs(
    x = "Percentage of first preference votes per division",
    y = "State",
    title = "First preference votes for the Greens party"
  )
```




]
.w-50[
```{r aus-election-plot2, fig.height = 5}
tdf3 %>%
  mutate(State = fct_reorder(State, perc_GRN)) %>%
  ggplot(aes(perc_GRN, State)) +
  ggbeeswarm::geom_quasirandom(groupOnX = FALSE, varwidth = TRUE) +
  labs(
    x = "Percentage of first preference votes per division",
    y = "State",
    title = "First preference votes for the Greens party"
  )
```
]

]

]
.panel[.panel-name[data]
.scroll-sign[.s400.f4[
```{r, ref.label = c("aus-election-data1", "aus-election-data3"), echo = TRUE}
```
]]]
.panel[.panel-name[R]
.f4[
```{r, ref.label=paste0("aus-election-plot", 1:2), echo = TRUE, eval = FALSE}
```
]]
]

--
<br><br>
<center>Sorting levels is (almost) always better when plotting</center>

---

class:  middle

# Order nominal variables meaningfully

<i class="fas fa-code"></i> **Coding tip**: use below functions to easily change the order of factor levels

```{r, eval = FALSE, echo = TRUE}
stats::reorder(factor, value, mean)
forcats::fct_reorder(factor, value, median)
forcats::fct_reorder2(factor, value1, value2, func)
```






---

# .orange[Case study] .circle.bg-orange.white[6] Aspirin use after heart attack

```{r meta-data, include = FALSE}
data("Fleiss93", package = "meta")
df6 <- Fleiss93 %>%
  mutate(total = n.e + n.c)
skimr::skim(df6)
```

.panelset[
.panel[.panel-name[📊]

.grid[
.item[
```{r meta-plot1, fig.height = 3}
df6 %>%
  mutate(study = fct_reorder(study, desc(total))) %>%
  ggplot(aes(study, total)) +
  geom_col() +
  labs(x = "", y = "Frequency") +
  guides(x = guide_axis(n.dodge = 2))

```

```{r meta-plot2, fig.height = 3}
df6 %>%
  mutate(
    study = ifelse(total < 2000, "Other", study),
    study = fct_reorder(study, desc(total))
  ) %>%
  ggplot(aes(study, total)) +
  geom_col() +
  labs(x = "", y = "Frequency")
```

]
.item[

* Meta-analysis is a statistical analysis that combines the results of multiple scientific studies.
* This data studies the use of aspirin for death prevention after myocardial infarction, or in plain terms, a heart attack.
* The ISIS-2 study has more patients than all other studies combined.
* You could consider lumping the categories with low frequencies together.
]


]

]
.panel[.panel-name[data]
.h300.f4.scroll-sign[
```{r meta-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r , ref.label=paste0("meta-plot", 1:2), echo = TRUE, eval = FALSE}
```
]]
]

.f5.footnote[
Fleiss JL (1993): The statistical basis of meta-analysis. *Statistical Methods in Medical Research* **2** 121–145<br>
Balduzzi S, Rücker G, Schwarzer G (2019), How to perform a meta-analysis with R: a practical tutorial, Evidence-Based Mental Health.
]

---

class: nostripheader middle


# Consider combining factor levels with low frequencies

<i class="fas fa-code"></i> **Coding tip**: the following family of functions help to easily lump factor levels together:

```{r fct_lump, echo = TRUE, eval = FALSE}
forcats::fct_lump()
forcats::fct_lump_lowfreq()
forcats::fct_lump_min()
forcats::fct_lump_n()
forcats::fct_lump_prop()
# if conditioned on another variable
ifelse(cond, "Other", factor)
dplyr::case_when(
  cond1 ~ "level1",
  cond2 ~ "level2",
  TRUE ~ "Other"
)
```


---

# .orange[Case study] .circle.bg-orange.white[7] Anorexia

```{r anorexia-data, include = FALSE}
data(anorexia, package = "MASS")

df9tab <- table(anorexia$Treat) %>%
  as.data.frame() %>%
  rename(Treatment = Var1, Frequency = Freq)

skimr::skim(anorexia)
```

.panelset[
.panel[.panel-name[📊]

.grid[
.item[
```{r anorexia-plot1}
ggplot(anorexia, aes(Treat)) +
  geom_bar() +
  labs(x = "", y = "Frequency")

```

```{r anorexia-table}
df9tab %>%
  knitr::kable() %>%
  kableExtra::kable_classic(full_width = FALSE)
```

]
.item[

**Table or Plot?**

{{content}}

]


]

]
.panel[.panel-name[data]
.h200.f4.scroll-sign[
```{r anorexia-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r , ref.label="anorexia-plot1", echo = TRUE, eval = FALSE}
```
```{r anorexia-plot2, echo = TRUE, eval = FALSE}
ggplot(anorexia, aes(Treat)) +
  stat_count(geom = "point", size = 4) +
  stat_count(geom = "line", group = 1) +
  labs(y = "Frequency", x = "")
```

]]
]

.f6.footnote[

Hand, D. J., Daly, F., McConway, K., Lunn, D. and Ostrowski, E. eds (1993) A Handbook of Small Data Sets. Chapman & Hall, Data set 285 (p. 229) <br>
Venables, W. N. & Ripley, B. D. (2002) Modern Applied Statistics with S. Fourth Edition. Springer, New York. ISBN 0-387-95457-0

]

--

* Table for accuracy, plot for visual communication

{{content}}

--

**Why not a point or line?**

```{r anorexia-plot2, fig.height = 2.5}
```

{{content}}

--

<div class="f4">
<ul>
<li>This can be appropriate depending on what you want to communicate </li>
<li>A barplot occupies more area compared to a point and the area does a better job of communicating size</li>
<li>A line is suggestive of a trend </li>
</ul>
</div>

---


# .orange[Case study] .circle.bg-orange.white[8] Titanic

```{r titanic-data, include = FALSE}
df9 <- as_tibble(Titanic)
skimr::skim(df9)
```

.panelset[
.panel[.panel-name[📊]

.flex[
.w-50[
.flex[
.w-50[
```{r titanic-plot1, fig.height = 2, fig.width = 4}
df9 %>%
  group_by(Class) %>%
  summarise(total = sum(n)) %>%
  ggplot(aes(Class, total)) +
  geom_col(fill = "#ee64a4") +
  labs(x = "", y = "Frequency")

```

```{r titanic-plot2, fig.height = 2, fig.width = 3}
df9 %>%
  group_by(Sex) %>%
  summarise(total = sum(n)) %>%
  ggplot(aes(Sex, total)) +
  geom_col(fill = "#746FB2") +
  labs(x = "", y = "Frequency")

```

]
.w-50[
```{r titanic-plot3, fig.height = 2, fig.width = 3}
df9 %>%
  group_by(Age) %>%
  summarise(total = sum(n)) %>%
  ggplot(aes(Age, total)) +
  geom_col(fill = "#C8008F") +
  labs(x = "", y = "Frequency")

```

```{r titanic-plot4, fig.height = 2, fig.width = 3}
df9 %>%
  group_by(Survived) %>%
  summarise(total = sum(n)) %>%
  ggplot(aes(Survived, total)) +
  geom_col(fill = "#795549") +
  labs(x = "Survived", y = "Frequency")

```

]
]


]
.w-50[

**What does the graphs for each categorical variable tell us?**

* There were more crews than 1st to 3rd class passengers
* There were far more males on ship; possibly because majority of crew members were male. You can further explore this by constructing two-way tables or graphs that consider both variables.
* Most passengers were adults. 
* More than two-thirds of passengers died. 

]


]

]
.panel[.panel-name[data]
.h350.f4.scroll-sign[
```{r titanic-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4.scroll-sign[.s500[
```{r , ref.label=paste0("titanic-plot", 1:4), echo = TRUE, eval = FALSE}
```
]]
]]

.f5.footnote[
British Board of Trade (1990), Report on the Loss of the ‘Titanic’ (S.S.). British Board of Trade Inquiry Report (reprint). Gloucester, UK: Allan Sutton Publishing
]


---


class: nostripheader middle

# Coloring bars


```{r, fig.width = 10}
g1 <- df9 %>%
  group_by(Class) %>%
  summarise(total = sum(n)) %>%
  ggplot(aes(Class, total)) +
  geom_col() +
  labs(x = "", y = "Frequency")

g2 <- df9 %>%
  group_by(Class) %>%
  summarise(total = sum(n)) %>%
  ggplot(aes(Class, total)) +
  geom_col(fill = "#006DAE") +
  guides(fill = FALSE) +
  labs(x = "", y = "Frequency") +
  scale_fill_discrete_qualitative()

g3 <- df9 %>%
  group_by(Class) %>%
  summarise(total = sum(n)) %>%
  ggplot(aes(Class, total)) +
  geom_col(aes(fill = Class)) +
  guides(fill = FALSE) +
  labs(x = "", y = "Frequency") +
  scale_fill_discrete_qualitative()

g4 <- df9 %>%
  group_by(Class) %>%
  summarise(total = sum(n)) %>%
  ggplot(aes(Class, total)) +
  geom_col(aes(fill = Class)) +
  guides(fill = FALSE) +
  scale_fill_manual("", values = c("#027C1E", "#649334", "#97A753", "#B9534C")) + # colorspace: terrain2 and heat
  labs(x = "", y = "Frequency") 

g1 + g2 + g3 + g4 + plot_layout(ncol=4)
```

--

* Colour here doesn't add information as the x-axis already tells us about the categories, but colouring bars can make it more visually appealing, and provide subtle clues. 
* If you have too many categories colour won't work well to differentiate the categories. 

---

# .orange[Case study] .circle.bg-orange.white[9] Opinion poll in Ireland Aug 2013

```{r poll-data, include = FALSE}
df9 <- tibble(
  party = c(
    "Fine Gael", "Labour", "Fianna Fail",
    "Sinn Fein", "Indeps", "Green", "Undecided"
  ),
  nos = c(181, 51, 171, 119, 91, 4, 368)
)
df9v2 <- df9 %>% filter(party != "Undecided")
df9
```

.panelset[
.panel[.panel-name[📊]

.grid[
.item[
```{r poll-plot1, fig.width = 7}
g9 <- df9 %>%
  ggplot(aes("", nos, fill = party)) +
  geom_col(color = "black") +
  labs(y = "", x = "") +
  coord_polar("y") +
  theme(
    axis.line = element_blank(),
    axis.line.y = element_blank(),
    axis.text = element_blank(),
    panel.grid.major = element_blank()
  ) +
  scale_fill_discrete_qualitative(name = "Party")
g9
```

```{r poll-plot2, fig.width = 7}
g9 %+% df9v2 +
  # below is needed to keep the same color scheme as before
  scale_fill_manual(values = qualitative_hcl(7)[1:6])

```

]
.item[

* Pie chart is popular in mainstream media but are not generally recommended as people are generally poor at comparing angles.
* 3D pie charts should definitely be avoided!
* Here you can see that there are many people that are "Undecided" for which political party to support and failing to account for this paints a different picture. 

]


]

]
.panel[.panel-name[data]
.f4[
```{r poll-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r , ref.label=paste0("poll-plot", 1:2), echo = TRUE, eval = FALSE}
```
]]]




---

class: middle

# Piechart is a stacked barplot just with a transformed coordinate system

--

```{r barplot, echo = TRUE, fig.height = 2, fig.width = 3}
df <- data.frame(var = c("A", "B", "C"), perc = c(40, 40, 20))
g <- ggplot(df, aes("", perc, fill = var)) +
  geom_col()
g
```
--

```{r piechart, echo = TRUE, fig.height = 2, fig.width = 3}
g + coord_polar("y")
```

---

class: middle

# Roseplot is a barplot just with a transformed coordinate system

--

```{r nonstacked-barplot, echo = TRUE, fig.height = 2, fig.width = 10}
dummy <- data.frame(
  var = LETTERS[1:20],
  n = round(rexp(20, 1 / 100))
)
g <- ggplot(dummy, aes(var, n)) +
  geom_col(fill = "pink", color = "black")
g
```
--

```{r roseplot, echo = TRUE, fig.height = 2, fig.width = 3}
g + coord_polar("x") + theme_void()
```

---
# Visual inference

.flex[

.item.w-50[

Typical plot description:


```{r eval=FALSE, echo=TRUE}
ggplot(data, aes(x=var1)) +
  geom_col()

ggplot(data, aes(x=var1)) +
  geom_bar()
```

<br><br>
*Is the distribution consistent with a sample from a binomial distribution with a given p?*

]
.item.w-5[
.white[space]
]
.item.w-45[

Potential simulation method from binomial

```{r eval=FALSE, echo=TRUE}
# Only one option
null_dist("var1", "binom", 
  list(size=n, p=phat))
```
]
]

---
# Lineup of tuberculosis count between sexes

.panelset[
.panel[.panel-name[📊]
```{r tb-lineup, fig.width=10, fig.height=4, out.width="80%"}
library(nullabor)
set.seed(252)
tb_oz_2012 <- tb %>%
  filter(iso3 == "AUS",
         year == 2012) %>%
  select(iso3, year, new_sp_m04:new_ep_m65) %>%
  pivot_longer(cols=new_sp_m04:new_ep_m65, 
               names_to = "var", 
               values_to = "count") %>%
  separate(var, into=c("new", "type", "sexage")) %>%
  select(-new) %>%
  filter(!(sexage %in% c("sexunk014", "sexunk04", 
                         "sexunk15plus", "sexunk514",
                         "f15plus", "m15plus"))) %>%
  group_by(sexage) %>%
  summarise(count = sum(count, na.rm=TRUE)) %>%
  mutate(sex=str_sub(sexage, 1, 1),
         age=str_sub(sexage, 2, str_length(sexage))) %>%
  group_by(sex) %>%
  summarise(count=sum(count)) %>%
  ungroup() %>%
  mutate(sex01 = ifelse(sex=="m", 0, 1)) %>%
  select(-sex)

ggplot(lineup(null_dist("count", "binom", 
                        list(size=sum(tb_oz_2012$count),
                               p=0.5)), 
              tb_oz_2012, n=10),
       aes(x=sex01, y=count)) +
  geom_col() +
  facet_wrap(~.sample, ncol=5) +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid.major = element_blank())
```

.f5[**Note**: This is nothing more than you can learn from a conventional hypothesis test of $$H_0: p=0.5$$. .monash-orange2[**Stay tuned for more interesting visual inference lineups in coming weeks!**]]
]
.panel[.panel-name[R]
.scroll-sign[.s500.f5[
```{r ref.label="tb-lineup", eval=FALSE, echo=TRUE}
```
]]
]
]

---

# Take away messages

.flex[
.w-70.f2[



<ul class="fa-ul">
{{content}}
</ul>


]
]

--

<li><span class="fa-li"><i class="fas fa-paper-plane"></i></span>Again, be prepared to do multiple plots</li>
{{content}}
--

<li><span class="fa-li"><i class="fas fa-paper-plane"></i></span>Changing bins or binwidth/bandwidth in histogram, violin or density plots can paint a different picture</li>
{{content}}
--

<li><span class="fa-li"><i class="fas fa-paper-plane"></i></span>Consider different representations of categorical variables (reordering meaningfully, lumping low frequencies together, plot or table, pie or barplot, missing categories)</li>

---
# Resources and Acknowledgement

- Slides originally written by Emi Tanaka and constructed with [`xaringan`](https://github.com/yihui/xaringan), [remark.js](https://remarkjs.com), [`knitr`](http://yihui.name/knitr), and [R Markdown](https://rmarkdown.rstudio.com).

---

```{r endslide, child="assets/endslide.Rmd"}
```
